How kids code and how we know it:\\An exploratory study in the Scratch project repository

### 2016-04-18 Paper/paper.tex
Section Results: Size and Complexity

\newcommand{\nPrograms}{250.166}
\newcommand{\nemptyPrograms}{14.285}
\newcommand{\nScriptPrograms}{233.491}
\newcommand{\nProgramsWithClones}{67.177}
\newcommand{\dup}{\emph{Duplication}~}

@@

\subsection{Program size and complexity}
Size -per Project:sprites, blocks and locs, events/triggers
lines in stage/sprites/procedures
nestendness	
cyclomatic complexity

lines do not count the ending line of enclosures
only sprites with code are counted
indentation: 1 for top-level, for nested (rest-1)/2 (goes 1,3,5, etc), for others normal

\subsubsection{Empty programs}

Also, interestingly enough, \nemptyPrograms~were empty (5.7\%), at least in terms of scripts. So Scratch users were sharing program already that did not contain code yet. In some cases \todo{analyze too?} they contained scripts and costumes but no code and in others they were entirety empty apart from the Scratch cat added by default.

>>>>

\newcommand{\nPrograms}{250,166}
\newcommand{\nAnalyzedPrograms}{247,798}
\newcommand{\nemptyPrograms}{14,307}
\newcommand{\nScriptPrograms}{233,491}
\newcommand{\nProgramsWithClones}{67,177}
\newcommand{\nLOC}{36,085,654}
\newcommand{\nscripts}{4,049,356}
\newcommand{\dup}{\emph{Duplication}~}

@@

In the following sections we describe the results obtained for each of the research questions through the analysis of the \nAnalyzedPrograms Scratch projects in our dataset.

\subsection{Program size and complexity}

The dataset is found to contain a relatively small number of projects without any code: \nemptyPrograms (5.77\%). Through random sampling we found that in some cases those projects contained sprites and costumes but no code and in others they were entirety empty apart from the Scratch cat added by default. Since those projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.

\begin{table*}[ht]
	\centering
	\begin{tabular}{lrrrrrr}
		\hline
		&\textbf{mean}&\textbf{min}&\textbf{Q1}&\textbf{median}&\textbf{Q3}&\textbf{max}\\
		\hline
		Sprites with code per project&5.68&1&1&2&5&525\\
		Scripts per project&17.35&1&2&5&12&3038\\
		Maximum indentation per project&4.56&1&3&3&5&488\\
		Lines of code (LOC) per project&154.55&1&12&29&76&34622\\
		LOC in Stage&4.80&0&0&0&3&2613\\
		LOC in Sprites&115.57&0&10&26&68&34613\\
		LOC in Procedures&34.17&0&0&0&0&20552\\
		McCabe Cyclomatic Complexity per script&1.58&0&0&0&0&246\\
		\hline
	\end{tabular}
	\caption{Size-related statistics from the dataset of \nScriptPrograms  non-empty Scratch projects}
	\label{tbl-size}
\end{table*}

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1sprites}
	\caption{Histogram of the number of sprites in the analyzed projects}
	\label{fig:size-sprites}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1scripts}
	\caption{Histogram of the number of scripts in the analyzed projects}
	\label{fig:size-scripts}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1locs}
	\caption{Histogram of the lines of code in the analyzed projects}
	\label{fig:size-locs}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1indent}
	\caption{Histogram of the maximum code indentation level in the analyzed projects}
	\label{fig:size-indent}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/3cyclomatic}
	\caption{Histogram of the McCabe cyclomatic complexity of the \nscripts analyzed scripts}
	\label{fig:cyclomatic}
\end{figure}

In Table \ref{tbl-size} we summarize statistics for metrics related to size and complexity. We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts per project, where each script is a code-block of interconnected Scratch blocks. For example, in Figure \ref{fig:ui} the cat sprite contains 2 scripts. We further measure the lines of code per project, considering each Scratch block to represent exactly one line of code, thus not counting the closing constructs of loops or conditional statements as separate lines. The maximum code indentation level is used as a code complexity metric. It was calculated using the JSON array depth? \todo{felienne, how did we measure this?}. This corresponds to code indentation level returning 1 for all first-level blocks and increasing by 2 for block enclosures in loops or conditional statements. Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and \ref{fig:size-indent} plot the distribution of those size metrics.

We find that the majority of Scratch projects are small; 75\% of the projects have up to 5 sprites, 12 scripts and 76 lines of code, while one fourth of the projects have up to 12 lines of code. On the other end, 5\% of the projects (11,712) are found to have more than 18 sprites and 4.8\% (11,214) consist of more than 500 lines of code. The analysis also highlighted some surprisingly large projects: 135 with more than 300 sprites and even 30 projects with more than 20,000 lines of code\footnote{The Scratch identifiers of those projects can be found in \todo{link}}.

The lines of code metric was further analyzed to understand code placement. As shown in Table \ref{tbl-size}, the majority of Scratch code---74.78\% out of \nLOC lines of code---is written within sprites. An additional 3.1\% of the total lines are found in the stage class. More interestingly, the remaining 22.11\% are lines within defined procedures, which are found in only 7.7\% (17,979) of the projects.

As a complexity metric we also used the McCabe cyclomatic complexity \cite{mcCabe76},  a quantitative measure of the number of linearly independent paths through a program's source code. This is calculated per script by counting the number of decision points in the script plus one. In Scratch, decision points can be the \texttt{if} and \texttt{if else} blocks. The results of the cyclomatic complexity metric per script are plotted in Figure \ref{fig:cyclomatic}. The majority (78.33\%) of \nscripts scripts contain no decision points, while 13.08\% have a cyclomatic complexity of 2, containing exactly one decision point. The complexity is higher, over 4, for 3.67\% of the scripts. The analysis also highlighted 209 scripts with cyclomatic complexity over 100 and up to 246.
